---
title: "Exploratory data analysis"
author: Lucien Bastin
date: 2025-09-13
format: html
knitr:
    opts_chunk:
        message: FALSE
        warning: FALSE
embed-resources: true
lightbox: true
toc: true
toc-location: right-body
---

# Load libraries and data

```{r}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(here)
library(tibble)
library(tidyr)
library(vegan)
```

```{r}
load(here("NEwR-2ed_code_data/NEwR2-Data/Doubs.RData"))
```

# First contact

```{r}
colnames(spe)
rownames(spe)
summary(spe)
range(spe)
apply(spe, 2, range)
```

Barplot of the distribution, all species confounded

```{r}
ab <- table(unlist(spe))

ggplot(as.data.frame(ab), aes(x = Var1, y = Freq, fill = Var1)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = gray(5:0 / 5), guide = "none") +
    labs(x = "Abundance class", y = "Frequency") +
    theme_bw()
```

# A closer look

```{r}
ggplot(spa, aes(x = X, y = Y)) +
    geom_path(color = "lightblue") +
    geom_text(aes(label = rownames(spa)), color = "red") +
    labs(x = "x coordinate (km)", y = "y coordinate (km)") +
    annotate("text", x = 68, y = 20, label = "Upstream", size = 5) +
    annotate("text", x = 15, y = 35, label = "Downstream", size = 5) +
    theme_test()
```

## Maps of some fish species

```{r}
spe_four <- spe |>
    select(Satr, Thth) |>
    rownames_to_column(var = "site") |>
    mutate(X = spa$X, Y = spa$Y) |>
    pivot_longer(
        cols = c(Satr, Thth),
        names_to = "species",
        values_to = "abundance"
    )

ggplot(spe_four, aes(x = X, y = Y)) +
    geom_path(color = "lightblue") +
    geom_point(aes(size = abundance), shape = 1, color = "red") +
    scale_size_continuous(range = c(-1, 8)) +
    labs(x = "x coordinate (km)", y = "y coordinate (km)") +
    facet_wrap(vars(species)) +
    theme_test()
```

## Species richness along the river

```{r}
site_richness <- apply(spe > 0, 1, sum)
site_richness_df <- spa |>
    mutate(richness = site_richness)

ggplot(site_richness_df, aes(x = X, y = Y)) +
    geom_path(color = "lightblue") +
    geom_point(aes(size = richness), color = "red") +
    scale_size_continuous(range = c(-1, 10)) +
    labs(x = "x coordinate (km)", y = "y coordinate (km)") +
    theme_test()
```

# Ecological data transformation

```{r}
spe_five <- spe[, 1:5]

plot_abundance <- function(data) {
    data |>
        rownames_to_column(var = "site") |>
        mutate(site = as.integer(site)) |>
        pivot_longer(
            cols = 2:6,
            names_to = "species",
            values_to = "abundance"
        ) |>
        ggplot(aes(x = site, y = abundance, fill = species)) +
        geom_bar(stat = "identity") +
        scale_x_continuous(breaks = 1:30) +
        theme_bw()
}

# Raw abundance
plot_abundance(spe_five)

# Presence-absence
plot_abundance(decostand(spe_five, method = "pa"))

# Standardization by maximum of each species
plot_abundance(decostand(spe_five, method = "max"))

# Standardization by species totals
plot_abundance(decostand(spe_five, "total", MARGIN = 2))

# Standardization by rows
plot_abundance(decostand(spe_five, "total"))

# Standardization by rows - Chord transformation
plot_abundance(decostand(spe_five, "normalize"))

# Square-root of relative abundance per site - Hellinger transformation
plot_abundance(decostand(spe_five, "hellinger"))

# Double standardization by columns and rows - Chi-square transformation
plot_abundance(decostand(spe_five, "chi.square"))

# Winsconsin standardization
plot_abundance(wisconsin(spe_five))
```

# Environmental data

```{r}
summary(env)
```

```{r}
env_long <- env |>
    select(dfs, ele, dis, oxy, nit) |>
    rownames_to_column("site") |>
    mutate(X = spa$X, Y = spa$Y) |>
    pivot_longer(ele:nit, names_to = "variable", values_to = "value") |>
    mutate(
        variable = factor(variable, levels = c("ele", "dis", "oxy", "nit"))
    ) |>
    group_by(variable) |>
    mutate(scaled_value = value / max(value))
```

```{r}
ggplot(env_long, aes(x = X, y = Y)) +
    geom_path() +
    geom_point(aes(color = variable, size = scaled_value)) +
    scale_size_continuous(range = c(-1, 5)) +
    facet_wrap(vars(variable)) +
    theme_test()
```

Variation of some descriptors along the river.

```{r}
ggplot(env_long, aes(x = dfs, y = value, color = variable)) +
    geom_line(show.legend = FALSE) +
    labs(x = "Distance from source (km)", y = element_blank()) +
    facet_wrap(
        vars(variable),
        scales = "free",
        strip.position = "left",
        labeller = as_labeller(
            c(
                ele = "Elevation (m)",
                dis = "Discharge (m³/s)",
                oxy = "Oxygen (mg/L)",
                nit = "Nitrate (mg/L)"
            )
        )
    ) +
    theme_test() +
    theme(strip.background = element_blank(), strip.placement = 'outside')
```

Bivariate relationships among environmental variables.

```{r}
#| fig-width: 12
#| fig-height: 10

panel.hist <- function(x, no.col = FALSE, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5))
  his <- hist(x, plot = FALSE)
  breaks <- his$breaks
  nB <- length(breaks)
  y <- his$counts
  y <- y / max(y)
  if (no.col)
    rect(breaks[-nB], 0, breaks[-1], y, col = "gray", ...)
  else
    rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}

pairs(
    env,
    panel = panel.smooth,
    diag.panel = panel.hist
)
```

Log-transformations of the variables.

```{r}
slope_df <- data.frame(slo = env$slo)

range(slope_df$slo)
```

```{r}
slope_hist_plot <- ggplot(slope_df, aes(x = slo)) +
    geom_histogram(
        binwidth = 10,
        breaks = seq(0, 50, 10),
        fill = "orange",
        color = "black"
    ) +
    labs(x = "Slope", y = "Frequency") +
    theme_test()

ln_slope_hist_plot <- ggplot(slope_df, aes(x = log(slo))) +
    geom_histogram(
        binwidth = 1,
        breaks = seq(-2, 4, 1),
        fill = "limegreen",
        color = "black"
    ) +
    labs(x = "ln(slope)", y = "Frequency") +
    theme_test()

slope_box_plot <- ggplot(slope_df, aes(y = slo)) +
    geom_boxplot(fill = "orange", width = 0.6) +
    xlim(-1,1) +
    labs(y = "Slope") +
    theme_test()

ln_slope_box_plot <- ggplot(slope_df, aes(y = log(slo))) +
    geom_boxplot(fill = "limegreen", width = 0.6) +
    xlim(-1,1) +
    labs(y = "ln(slope)") +
    theme_test()

ggarrange(
    slope_hist_plot,
    ln_slope_hist_plot,
    slope_box_plot,
    ln_slope_box_plot,
    nrow = 2,
    ncol = 2
)
```

Standardization of all environmental variables.

```{r}
env_z <- decostand(env, "standardize") # z-scores, mean = 0 and sd = 1
```